|

### 2021-10-13 manuscripts/cprd/manuscript.Rmd
Pre-final changes to manuscript

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE) # Remove to knit with plots/data
options(scipen = 999)
source(here::here("R/doc_options.R"))
source(here::here("R/helper.R"))

@@

knitr::read_chunk(here::here("R/04-Code-CPRD-Analysis.R"))
doc_type <- knitr::opts_knit$get('rmarkdown.pandoc.to')
```

@@

__Data/code availability__
This analysis used the CPRD-GOLD primary care dataset March 2016 snapshot (ISAC 15_246R), which is available upon application to the CPRD Independent Scientific Advisory Committee. The code lists used to define the outcomes and covariates for this study, in addition to the cleaning and analysis scripts used to create the study cohort and perform the analyses, are available on GitHub (https://github.com/mcguinlu/CPRD-LRA), and were archived at the time of submission on Zenodo (DOI: __TBC__).

@@

<!-- TODO Check scale in figures -->
<!-- TODO Check cross link between supplementary tables and text -->

@@

__Check TODOs!__

<!-- TODO Fix total length of follow-up between text and table 2 -->

__Background:__ There is some evidence that circulating blood lipids play a role in the development of Alzheimer's disease (AD) and dementia. As a result, these modifiable risk factors could be targeted by existing lipid-regulating agents, including statins, for the prevention of dementia. Here, we test the association between lipid-regulating agents and risk of dementia and related conditions in the Clinical Practice Research Datalink (CPRD), an electronic health record database.

__Methods:__ A cohort study using data from the CPRD, routinely collected between January 1995 and March 2016, was performed. Multivariable Cox proportional hazard models, allowing for a time-varying treatment indicator, were used to estimate the association between seven lipid-regulating drug classes (vs. no drug) and five dementia outcomes (all-cause, vascular and other dementia, and probable and possible Alzheimer's disease).

__Results:__ We analyzed `r totaltext` with a total follow-up of `r comma(total_followup)` participant-years (median: `r fu_text`). We found little evidence that lipid-regulating agents were associated with Alzheimer's disease (probable `r probad_text`; possible `r possad_text`), but there was evidence of an increased risk of all-cause (`r anydem_text`), vascular (`r vasdem_text`) and other dementia (`r othdem_text`). Evidence from a number of control outcomes (ischaemic heart disease `r ihd_text`; backpain `r backpain_text`; and Type 2 diabetes `r dm_type2_text`) indicated the presence of substantial residual confounding by indication.

__Conclusion:__ We found little evidence that lipid-regulating agents were associated with reduced Alzheimer’s disease risk. There was some evidence of an increased the risk of all-cause, vascular and other dementia, likely due to residual confounding by indication.

@@

* A large cohort of patients from the Clinical Practice Research Datalink electronic health record database was assembled to examine the effect of lipid-regulating agents, such as statins, on dementia outcomes.

* Little evidence that lipid-regulating agents were associated with Alzheimer’s disease, but there was some evidence for a harmful effect on all-cause, vascular and other dementia. In all cases, the estimated associations were driven by the any statin subgroup, which comprised most participants in our cohort.

@@

Several cardiovascular factors have been identified as potential risk factors for dementia,[@walker2020] and of these, circulating lipid levels represent a promising target for intervention due to the ready availability of lipid-modifying treatments. In this context, determining whether lipid-regulating agents (LRA) could be repurposed for the prevention of dementia and related diseases would be helpful in the development of evidence-based prevention policy. Several existing prospective studies have examined the association of LRA use with dementia.[@chu2018; @larsson2018; @ancelin2012; @bettermann2012;@rea2005] However, many of these studies are small, record few outcomes, and have limited follow-up.

@@

We therefore aimed to test the association between several major classes of LRA and all-cause dementia, Alzheimer's disease, vascular dementia and other dementia, in the Clinical Practice Research Datalink (CPRD), a large, population-based electronic health record (EHR) database.[@herrett2015]

@@

We performed a cohort study using data from the CPRD. Our initial sample included all participants registered at a participating practice between 1 January 1995 and 29 February 2016 who had a flag for "research quality" data. Records pre-dating the 1995 cut-off were excluded from the analysis as data quality and reliability is thought to be higher after this date.[@wolf2019] All events of interest were identified using predetermined code lists, which are available for inspection (see [Data/code availability](#data-code-avail)).

@@

Participants were included in our study cohort if their record contained any of the following index events: a Read code for a diagnosis of hypercholesterolemia or related condition; a Read code for prescription of a LRA (such as statins); a total cholesterol test result of >4mmol/L; or a low density lipoprotein cholesterol (LDL-c) test result of >2mmol/L.

These index events allowed us to define a population of participants who were either at risk of hypercholesterolemia, as indicated by the elevated total or LDL-c test results, or had already been diagnosed with it, as indicated by a diagnostic code or related prescription. This approach, conditioning entry into the study on being either “at-risk” or already diagnosed with hypercholesterolemia, was employed to reduce confounding by indication that we would expect to observe in a general population cohort.

The index date for a participant was defined as the date where the first relevant code or test result (as detailed above) was recorded on their clinical record. Participants were followed up until the earliest of: an outcome of interest; death; end of follow-up (29 February 2016); last registration date with their GP practice; or the last CPRD collection date for their practice. Participants were removed from our sample if they were less than 40 years of age, had less than 12 months of "research quality" data prior to their index date, were simultaneously prescribed more than one LRA (due to the difficulty of assigning these patients to a single exposure group), or were diagnosed with dementia before or on the date of the index event.

@@

We considered five outcomes as part of this analysis: probable  Alzheimer’s disease, possible Alzheimer’s disease, vascular dementia, other dementia, and a composite all-cause dementia outcome. When two or more outcomes were coded in a participant’s clinical record, a decision tree was used to differentiate between them (Supplementary Figure 1). The diagnosis date of the outcome was determined by the first record of a relevant code.

@@

The analysis was adjusted for a range of baseline covariates including sex, grouped year of entry into the cohort (<=2000, 2001-2005, 2006-2010, >2010), Charlson co-morbidity index, Index of Multiple Deprivation (IMD), consultation rate, alcohol (current, former,  never), smoking (current, former,  never), BMI, baseline total cholesterol, and history of cardiovascular disease, coronary bypass surgery, coronary artery disease, peripheral arterial disease, hypertension, chronic kidney disease, and Type 1 and Type 2 diabetes. All covariates were determined at index and definitions for each can be found in Supplementary Table 2.

@@

All analyses were performed in STATA 15. Cox proportional hazard models with a time-varying treatment indicator were used to estimate the hazard ratio and corresponding 95% confidence intervals, allowing for potential clustering of outcomes by practice. Participant's age was used as the time axis for all models.[@lamarca1998; @gail2009; @pencina2007] To observe the effect of adjusting for additional covariates, we compared models adjusted for age only and age and sex to the fully adjusted model. Additional analyses stratified by outcome and drug class were also performed.

@@

We performed several sensitivity analyses. As statins are contraindicated in pregnancy,[@nationalinstituteforhealthandcareexcellence]  we ran the models described above but excluding participants below the age of 55. Given the different ability of lipophilic statins to cross the blood-brain barrier,[@sierra2011] we further stratified the statin exposure group into lipophilic (Atorvastatin, Lovastatin, Simvastatin, Cerivastatin) and hydrophilic (Pravastatin, Rosuvastatin, Fluvastatin) statins. Finally, we included three control outcomes with known associations with statin use.[@collins2016a; @herrett2021] Using the fully adjusted model, we investigated the association between LRA and back pain (negative control), ischaemic heart disease (positive protective control), and Type 2 diabetes (positive harmful control).

@@

A total of `r format(attrition[14,1], big.mark = ",")` participants met the inclusion criteria for our cohort (see Supplementary Figure 2 for the attrition flowchart), with a total follow-up of `r comma(total_followup)` patient years at risk. Most participants were included in the cohort due to an elevated test result (`r index_event_text`). The median age at index was `r age_text` and participants were followed up for a median of `r fu_text`. During follow-up, an all-cause dementia diagnosis was recorded for `r format(p1[1,2], big.mark = ",")` patients  (`r format(p1[2,2], big.mark = ",")` probable Alzheimer's disease, `r format(p1[3,2], big.mark = ",")` possible Alzheimer's disease, `r format(p1[4,2], big.mark = ",")` vascular dementia, `r format(p1[5,2], big.mark = ",")` other dementia). The distribution of baseline characteristics across the drug classes can be seen in Table \@ref(tab:cprdCharacteristics-table).

Most participants (`r percentage.statins`) prescribed a lipid-regulating agent were prescribed a statin. We excluded the "Ezetimibe and statins" and "Nicotinic acid groups" classes from subsequent analysis based on the extremely small number of participants in these groups (Table \@ref(tab:cprdCharacteristics-table)). The stopping, addition and switching of drug classes was common across all exposure groups (Supplementary Table 3).

@@

Full covariate information was available for `r missingtext`. Five key variables had some missing data: IMD 2010 score, a proxy for socioeconomic position that is measured as twentiles with 1 indicating the least deprived and 20 indicating the most deprived, was missing for `r imdtext`, because it is only recorded for English practices; alcohol status was missing for `r alcoholtext`; smoking status was missing for `r smokingtext`; BMI, or a calculated BMI from height and weight measurements, was missing for `r bmitext`; baseline total cholesterol was missing for `r choltext`; and baseline LDL cholesterol was missing for `r ldltext`.

@@

As shown in Figure \@ref(fig:cprdPrimary), our results show little evidence of an effect of lipid-regulating agents on probable (`r probad_text`) and possible (`r possad_text`) Alzheimer's disease when compared with no treatment, except for fibrates on probable Alzheimer's disease (`r probad_fib_text`).

@@

In contrast to the findings for Alzheimer's disease outcomes, lipid-regulating agents were associated with an increased risk of a subsequent diagnosis of vascular dementia (`r vasdem_text`) or other dementia (`r othdem_text`). Again, this effect was driven mainly by the any statin subgroup, but there was some evidence that ezetimibe was associated with an increased risk of vascular (`r vasdem_eze_text`) and other (`r othdem_eze_text`) dementia.

@@

For the composite all-cause dementia outcome, we found treatment with a lipid-regulating agent was associated with a slightly increased risk (`r anydem_text`), but the magnitude of the association was not as extreme as that observed for the vascular dementia subgroup. There was also some evidence that fibrates were associated with increased risk of all-cause dementia (`r anydem_fib_text`).

@@

Adjustment for additional covariates beyond age and sex had a limited impact (Supplementary Figure 3), except for the Probable Alzheimer's disease outcome, where the full adjustment attenuated to the null the protective effect observed when adjusting only for age and sex.

Removing participants aged 55 and under at index from our analysis had minimal effect on our estimates (Supplementary Figure 4). When stratifying by statin properties, hydrophilic statins were less harmful in the any, vascular and other dementia outcomes compared to lipophilic statins (Supplementary Figure 5). Additionally, in the Alzheimer's disease outcomes, hydrophilic statins had a small protective effect, compared to the absence of evidence for an effect for lipophilic statins.

@@

There was little evidence that lipid-regulating agents had any observed effect on probable and possible Alzheimer's when compared with no treatment, but some evidence they were associated with an increased risk of an all-cause dementia, vascular dementia and other dementia diagnosis. The effect observed in each case was driven by the any statin subgroup, which included a substantial majority of participants. For the other drug classes, there was limited evidence of an association with any outcome, with two exceptions. Ezetimibe was associated with increased risk of vascular and other dementia, while fibrates were associated with increased risk of all-cause dementia and probable Alzheimer's disease.

@@

## Comparison to other literature

<!-- TODO Need a little bit here on comparison with Smeeth -->

@@

A recent Cochrane Review identified two randomized trials comparing treatment with statins versus non-treatment for the prevention of dementia, only one of which presented information on the incidence of dementia.[@mcguinness2016] This study (Heart Protection Study) showed no effect of treatment with simvastatin on all-cause dementia risk (OR: 1.00,  95%CI:0.61-1.65),[@heartprotectionstudycollaborativegroup2002] but concerns were raised over the diagnostic criteria used. A meta-analysis of 30 observational studies found a reduced risk of all-cause dementia was associated with statin treatment (RR 0.83, 95%CI: 0.79–0.87).[@poly2020] This conflicts with the findings of our analysis, where statin use was associated with an increased risk of all-cause dementia (`r anydem_text`). However, some of the included studies in the meta-analysis specifically exclude vascular dementia from the definition of all-cause dementia,[@chao2015] which may limit the ability for comparison with the results of this analysis.

Additionally, a previous analysis of the THIN EHR database using a propensity score matching analysis found a protective effect of statins on all-cause dementia (HR:0.81, 95%CI:0.69-0.96),[@smeeth2009]. Differences in dementia coding over time, in addition to a different set of covariates adjusted for in the analysis, may go someway to explaining the discrepancy.

@@

Our results are broadly in line with the findings of two distinct approaches examining the effect of statin treatment on subsequent Alzheimer's disease. No randomized trials of statins for the prevention of Alzheimer's disease have been reported, but a recent meta-analysis of 20 observational studies found statins were associated with a reduced risk of Alzheimer's disease (RR 0.69, 95% CI 0.60–0.80) with stronger evidence than observed in our analysis.[@poly2020] Additionally, a recent Mendelian randomization study examining the effect of genetic inhibition of HMGCR on Alzheimer’s disease (a genetic proxy for statin treatment) provided equivocal evidence (OR: 0.91, 95%CI: 0.63-1.31) but was not inconsistent with our results.[@williams2020]

Our additional analysis found no difference in effect between lipophilic and hydrophilic statins for the prevention of Alzheimer's disease, consistent with a recent meta-analysis of observational studies.[@chu2018]

<!-- TODO Need to cite Smeeth paper and discuss -->

@@

Far fewer studies have tested the association between lipid-regulating agents and vascular dementia or other dementia. A recent review found four observational studies examining the association of statins and vascular dementia found limited evidence for an effect (RR:0.93, 95% CI 0.74–1.16).[@poly2020] This contrasts with the increased effect found in our analysis (`r vasdem_text`). An additional analysis found that lipophilic statins were more harmful than hydrophilic statins in vascular dementia, potentially due to their ability to cross the blood brain barrier.

@@

Apart from statins, few studies examining a lipid-regulating agent have been reported. One of the few classes for which data was available were fibrates, minimal evidence of an effect on all-cause dementia was identified,[@ancelin2012] inconsistent with our finding of a small increase in all-cause dementia risk in those prescribed a fibrate.

A previous Mendelian randomization study provided an estimate of the effect of ezetimibe on Alzheimer's disease (OR: 1.17, 95%CI: 0.73-1.87),[@williams2020] consistent with the finding from our analyses. To our knowledge, there is no previous study of the effect of preventative treatment with ezetimibe on risk of vascular dementia, and so we cannot compare our unexpected finding for that comparison.

@@

## Strengths and Limitations

@@

However, the findings of our analysis are subject to several limitations. There was a  possibility of differential outcome misclassification based on the exposure, as we cannot exclude the possibility that for people with memory complaints, a diagnosis of vascular dementia might be made more frequently than Alzheimer’s disease if their medical records contain prescriptions for lipid-regulating agents.

Further, there is a potential for non-differential misclassification of the outcome based on the use of electronic health records to identify dementia cases.[@wilkinson2018; @mcguinness2019validity]

Our study is likely to be subject to confounding by indication, which could provide a potential explanation for the observed increased risk of vascular and other dementia with lipid regulating agent use. Supporting evidence for this interpretation comes from a variety of sources, including the results of the control outcome analyses. The slight harmful effect for the backpain outcome is substantially smaller than that observed for the ischaemic heart disease outcome, indicating that the majority of the uncontrolled confounding is likely related to vascular factors. Additionally, we obtained the expected harmful result for Type 2 diabetes, where statins' mechanism of action on this outcome is unlikely to be vascular.[@collins2016a; @macedo2014] Further supporting evidence comes from the increasingly harmful effect when moving from the Probable/Possible Alzheimer's disease to other dementia to vascular dementia outcomes, indicating that confounding by indication likely increases as the proportion of cases with a vascular component increases. A review of other available literature suggests that this observation (a harmful effect of lipid regulating agents on vascular-related outcome due to confounding by indication) is not unusual. Using a conventional epidemiological technique, a previous analysis also found an increased risk of coronary heart disease (analogous to the ischaemic heart disease outcome used in our analysis) in those taking statins (`r estimate(1.31,1.04, 1.66, type = "HR")`).[@danaei2013b] Following control for confounding by indication in that study through use of a trial emulation analysis, the expected protective effect of statins was observed.

Finally, there is also the potential for reverse causation in this analysis. Dementia and associated conditions have a long prodromal period, during which preclinical disease could cause indications for the prescription of a lipid-regulating agent.

@@

We have provided new evidence on the association of lipid-regulating agent prescription with all-cause dementia, Alzheimer’s disease, vascular dementia, and other dementia. We found limited evidence to support the use of lipid-regulating agents for the prevention of  probable or possible Alzheimer’s disease, but they were associated with an increased risk of all-cause, vascular and other dementias. In all cases, the estimated associations were driven by those observed in the any statin subgroup, which comprised the majority of participants in our cohort.

We have attempted to account for important sources of bias in our analysis and provide a comparison with other available literature. However, there is a strong potential for unmeasured confounding, misclassification and reverse causation, which may relate to the unexpected increase in risk of vascular dementia associated with statin use. Other analytical methods. Future research should aim to address these biases, potentially via a trial emulation analysis.

@@

# Main table

@@

<!----------------------------------------------------------------------------->
(ref:followUp-caption) Summary of person-time at risk and number of events by exposure group.

# Main figure

<!-- TODO 11,119 participants removed because their endate was . This explains the discrepancy between the total number in this figure and the table above. -->

>>>>

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE) # Remove to knit with plots/data
source(here::here("R/helper.R"))

@@

knitr::read_chunk(here::here("R/05-Code-CPRD-Analysis.R"))
doc_type <- knitr::opts_knit$get('rmarkdown.pandoc.to')
```

@@

@@

<!-- TODO Check cross link between supplementary tables and text -->

@@

__Background:__ There is some evidence that circulating blood lipids play a role in the development of Alzheimer's disease (AD) and dementia. These modifiable risk factors could be targeted by existing lipid-regulating agents, including statins, for the prevention of dementia. Here, we test the association between lipid-regulating agents and risk of dementia and related conditions in the Clinical Practice Research Datalink (CPRD), a United Kingdom electronic health record database.

__Methods:__ A retrospective cohort study was performed using routinely collected data from the CPRD (January 1995 and March 2016). Multivariable Cox proportional hazard models, allowing for a time-varying treatment indicator, were used to estimate the association between seven lipid-regulating drug classes (vs. no drug) and five dementia outcomes (all-cause, vascular and other dementias, and probable and possible Alzheimer's disease).

__Results:__ We analyzed `r totaltext` with a total follow-up of `r comma(total_followup)` patient-years (median: `r fu_text`). We found little evidence that lipid-regulating agents were associated with incidence of Alzheimer's disease (probable `r probad_text`; possible `r possad_text`), but there was evidence of an increased risk of all-cause (`r anydem_text`), vascular (`r vasdem_text`) and other dementias (`r othdem_text`). Evidence from a number of control outcomes indicated the presence of substantial residual confounding by indication (ischaemic heart disease `r ihd_text`; backpain `r backpain_text`; and Type 2 diabetes `r dm_type2_text`).

__Conclusion:__ Lipid-regulating agents were not associated with reduced Alzheimer’s disease risk. There was some evidence of an increased the risk of all-cause, vascular and other dementias, likely due to residual confounding by indication.

@@

* A large cohort of patients from the Clinical Practice Research Datalink electronic health record database was assembled to examine the association of lipid-regulating agents, such as statins, with dementia outcomes.

* Little evidence that lipid-regulating agents were associated with Alzheimer’s disease, but there was some evidence for a harmful association with all-cause, vascular and other dementias. In all cases, the estimated associations were driven by the any statin subgroup, which comprised most participants in our cohort.

@@

Several cardiovascular factors have been identified as potential risk factors for dementia,[@walker2020] and of these, circulating lipid levels represent a promising target for intervention due to the ready availability of lipid-modifying treatments. In this context, determining whether lipid-regulating agents (LRA) could be repurposed for the prevention of dementia and related diseases would be helpful in the development of evidence-based prevention policy. Several existing prospective studies have examined the association of LRA use with dementia.[@chu2018; @larsson2018; @ancelin2012; @bettermann2012;@rea2005] However, many of these studies are small, record few outcomes, and had limited follow-up.

@@

We therefore aimed to test the association between several major classes of LRA and all-cause dementia, Alzheimer's disease, vascular dementia and other dementias, in the Clinical Practice Research Datalink (CPRD), a large, population-based electronic health record (EHR) database from the United Kingdom.[@herrett2015]

@@

We defined a longitudinal cohort using data from the CPRD. Our initial sample included all participants registered at a participating practice between 1 January 1995 and 29 February 2016 who had a flag for "research quality" data. Records pre-dating the 1995 cut-off were excluded from the analysis as data quality and reliability are thought to be higher after this date.[@wolf2019] All events of interest were identified using predetermined code lists, which are available for inspection (see [Data/code availability](#data-code-avail)).

@@

Participants were included in our study cohort if their record contained any of the following index events: (a) a Read code for a diagnosis of hypercholesterolemia or related condition; (b) a Read code for prescription of a LRA (such as statins); (c) a total cholesterol test result of >4mmol/L; (d) or a low-density lipoprotein cholesterol (LDL-c) test result of >2mmol/L.

These index events allowed us to define a population of participants who were either at risk of hypercholesterolemia, as indicated by the elevated total or LDL-c test results, or had already been diagnosed with it, as indicated by a diagnostic code or related prescription. This approach, conditioning entry into the study on being either “at-risk” or already diagnosed with hypercholesterolemia, was employed to reduce the "confounding by indication" that we would expect to observe if we had used a general population cohort.

An index date for each participant was defined as the date when the first relevant code or test result (as detailed above) was recorded on their clinical record. Participants were followed up until the earliest of (a) an outcome of interest; (b) death; (c) end of follow-up (29 February 2016); (d) last registration date with their GP practice; or (e) the last CPRD collection date for their practice. Participants were removed from our sample if they were less than 40 years of age at entry (as these patients are less likely to be prescribed a LRA), had less than 12 months of "research quality" data prior to their index date, were simultaneously prescribed more than one LRA (due to the difficulty of assigning these patients to a single exposure group), or were diagnosed with dementia before or on the date of the index event.

@@

We considered five outcomes as part of this analysis: probable  Alzheimer’s disease, possible Alzheimer’s disease, vascular dementia, other dementias, and a composite all-cause dementia outcome. When two or more outcomes were coded in a participant’s clinical record, a decision tree was used to differentiate between them (Supplementary Figure 1). The diagnosis date of the outcome was determined by the first record of a relevant code.

@@

The analysis was adjusted for a range of baseline covariates including sex, grouped year of entry into the cohort (<=2000, 2001-2005, 2006-2010, >2010), Charlson co-morbidity index, Index of Multiple Deprivation (IMD), consultation rate, alcohol use (current, former,  never), smoking (current, former,  never), BMI, baseline total cholesterol, and history of cardiovascular disease, coronary bypass surgery, coronary artery disease, peripheral arterial disease, hypertension, chronic kidney disease, and Type 1 and Type 2 diabetes. These variables were selected as potential confounders between dementia outcomes and use of a LRA. All covariates were determined at the index date and definitions for each can be found in Supplementary Table 2.

@@

All analyses were performed in STATA (V16, StataCorp LLC, College Station, TX). Cox proportional hazard models with a time-varying treatment indicator were used to estimate the hazard ratio and corresponding 95% confidence intervals, allowing for potential clustering of outcomes by practice. Participant's age was used as the time axis for all models.[@lamarca1998; @gail2009; @pencina2007] To observe the effect of adjusting for additional covariates, we compared models adjusted for age only and age and sex with the fully adjusted model. Additional analyses stratified by outcome and drug class were also performed.

@@

We performed several sensitivity analyses. As statins are contraindicated in pregnancy,[@nationalinstituteforhealthandcareexcellence]  we ran the models described above but excluding participants below the age of 55. Given the different ability of lipophilic statins to cross the blood-brain barrier,[@sierra2011] we further stratified the statin exposure group into lipophilic (Atorvastatin, Lovastatin, Simvastatin, Cerivastatin) and hydrophilic (Pravastatin, Rosuvastatin, Fluvastatin) statins. Finally, we included three alternative outcomes with known associations with statin use as positive or negative controls using the fully adjusted model: back pain (negative control), ischaemic heart disease (positive protective control), and Type 2 diabetes (positive harmful control).[@collins2016a; @herrett2021]

@@

A total of `r format(attrition[14,1], big.mark = ",")` participants met the inclusion criteria for our cohort (see Supplementary Figure 2 for the attrition flowchart), with a total follow-up of `r comma(total_followup)` patient years at risk. Most participants were included in the cohort due to an elevated test result (`r index_event_text`). The median age at the index date was `r age_text` and participants were followed up for a median of `r fu_text`. During follow-up, an all-cause dementia diagnosis was recorded for `r format(p1[1,2], big.mark = ",")` patients  (`r format(p1[2,2], big.mark = ",")` probable Alzheimer's disease, `r format(p1[3,2], big.mark = ",")` possible Alzheimer's disease, `r format(p1[4,2], big.mark = ",")` vascular dementia, `r format(p1[5,2], big.mark = ",")` other dementia). The number of events, time-at-risk and crude rates for each drug class, tabulated by dementia outcome, are shown in Table \@ref(tab:followUp-table). Most participants (`r percentage.statins`) prescribed a lipid-regulating agent were prescribed a statin. We excluded the "Ezetimibe and statins" and "Nicotinic acid groups" classes from subsequent subgroup analyses based on the extremely small number of participants in these groups (Table \@ref(tab:followUp-table)). The distribution of baseline characteristics across the remaining seven drug classes can be seen in Table \@ref(tab:cprdCharacteristics-table). The stopping, addition and switching of drug classes was common across all exposure groups (Supplementary Table 3).

@@

Full covariate information was available for `r missingtext`. Five key variables had some missing data: IMD 2010 score, a proxy for socioeconomic position that is measured as twentiles with 1 indicating the least deprived and 20 indicating the most deprived, was missing for `r imdtext`, because it is only recorded for selected English practices; alcohol status was missing for `r alcoholtext`; smoking status was missing for `r smokingtext`; BMI, or a calculated BMI from height and weight measurements, was missing for `r bmitext`; baseline total cholesterol was missing for `r choltext`; and baseline LDL cholesterol was missing for `r ldltext`.

@@

As shown in Figure \@ref(fig:cprdPrimary), our results show little evidence of an association between lipid-regulating agents and probable (`r probad_text`) or possible (`r possad_text`) Alzheimer's disease when compared with no treatment, except for an adverse association between fibrates and probable Alzheimer's disease (`r probad_fib_text`).

@@

In contrast to the findings for Alzheimer's disease outcomes, lipid-regulating agents were associated with an increased risk of a subsequent diagnosis of vascular dementia (`r vasdem_text`) or other dementias (`r othdem_text`). Again, the observed estimate was driven mainly by the any statin subgroup, but there was some evidence that ezetimibe was associated with an increased risk of vascular (`r vasdem_eze_text`) and other (`r othdem_eze_text`) dementia.

@@

For the composite all-cause dementia outcome, we found treatment with a lipid-regulating agent was associated with a slightly increased risk (`r anydem_text`), which lies between the associations for the Alzheimer and non-Alzheimer dementia outcomes as would be expected. There was also some evidence that fibrates were associated with increased risk of all-cause dementia (`r anydem_fib_text`).

@@

Adjustment for additional covariates beyond age and sex had a limited impact (Supplementary Figure 3), except for the probable Alzheimer's disease outcome, where the full adjustment attenuated to the null the protective association observed when adjusting only for age and sex.

Removing participants aged 55 and under at the index date from our analysis had minimal effect on our estimates (Supplementary Figure 4). When stratifying by statin properties, hydrophilic statins were less harmful in relation to the any, vascular and other dementias outcomes compared with lipophilic statins (Supplementary Figure 5). Similarly, hydrophilic statins were associated with a reduced incidence of Alzheimer's disease, compared with the absence of evidence for an association with lipophilic statins.

@@

There was little evidence for an association between lipid-regulating agents and probable and possible Alzheimer's when compared with no treatment, but some evidence they were associated with an increased risk of an all-cause dementia, secondary to their association with vascular and other dementias diagnoses. The association observed in each case was driven by the any statin subgroup, which included a substantial majority of participants. For the other drug classes, there was limited evidence of an association with any outcome, with two exceptions. Ezetimibe was associated with increased risk of vascular and other dementias, while fibrates were associated with increased risk of all-cause dementia and probable Alzheimer's disease.

@@

## Comparison with other literature

@@

A recent Cochrane Review identified two randomized trials comparing treatment with statins versus non-treatment for the prevention of dementia, only one of which presented information on the incidence of dementia.[@mcguinness2016] This study (Heart Protection Study) showed no effect of treatment with simvastatin on all-cause dementia risk (OR: 1.00,  95%CI:0.61-1.65),[@heartprotectionstudycollaborativegroup2002] but concerns were raised over the diagnostic criteria used. A meta-analysis of 30 observational studies found a reduced risk of all-cause dementia was associated with statin treatment (RR 0.83, 95%CI: 0.79–0.87).[@poly2020] This conflicts with the findings of our analysis, where statin use was associated with an increased risk of all-cause dementia (`r anydem_text`). Some of the included studies in the meta-analysis specifically exclude vascular dementia from the definition of all-cause dementia,[@chao2015] which may limit the ability for comparison with our findings for the all-cause dementia outcome.

Additionally, a previous analysis of the THIN EHR database using a propensity-score matched analysis found a protective effect of statins on all-cause dementia (HR:0.81, 95%CI:0.69-0.96)[@smeeth2009]. Differences in the code-lists used to define dementia outcomes, in addition an analytical approach that adjusted for covariates defined after the index date, may go some way to explaining the discrepancy.

@@

Our results are broadly in line with the findings of two distinct approaches examining the effect of statin treatment on subsequent Alzheimer's disease. No randomized trials of statins for the prevention of Alzheimer's disease have been reported, but a recent meta-analysis of 20 observational studies found statins were associated with a reduced risk of Alzheimer's disease (RR 0.69, 95% CI 0.60–0.80) with stronger evidence than observed in our analysis.[@poly2020] This review included case-control studies and analyses likely to be at risk of immortal time bias, which may account for the discrepancy with our findings. Additionally, a recent Mendelian randomization study examining the effect of genetic inhibition of HMGCR on Alzheimer’s disease (a genetic proxy for statin treatment) provided equivocal evidence (OR: 0.91, 95%CI: 0.63-1.31) but was consistent with our results.[@williams2020]

Our additional analyses of statin properties found little evidence of differences in associations of lipophilic and hydrophilic statins and incidence of Alzheimer’s disease, consistent with a recent meta-analysis of observational studies.[@chu2018]

@@

Far fewer studies have tested the association between lipid-regulating agents and vascular dementia or other dementias. A recent review found four observational studies examining the association of statins and vascular dementia found limited evidence for an effect (RR:0.93, 95% CI 0.74–1.16).[@poly2020] This contrasts with the harmful association found in our analysis (`r vasdem_text`). When stratifying by lipid properties, lipophilic statins were more harmful than hydrophilic statins in vascular dementia, potentially due to their ability to cross the blood brain barrier.

@@

Apart from statins, few studies examining a lipid-regulating agent have been reported. One of the few classes for which a previous analysis was available were fibrates, which found little evidence of an association with all-cause dementia was identified,[@ancelin2012] inconsistent with our finding that patients prescribed fibrates had higher all-cause dementia risk than those prescribed other lipid lowering agents.

A previous Mendelian randomization study found little evidence that genetic variants that proxy for ezetimibe affect risk of Alzheimer’s disease (OR: 1.17, 95%CI: 0.73-1.87),[@williams2020] consistent with our findings. To our knowledge, there is no previous study of the effect of preventative treatment with ezetimibe on risk of vascular dementia.

@@

## Strengths and limitations

@@

However, the findings of our analysis are subject to several limitations. Confounding by indication is a major bias in pharmacoepidemiological studies and could provide a potential explanation for the observed increased risk of vascular and other dementias with lipid regulating agent use. Patients with vascular risk factors are more likely to receive a statin prescription and also to be diagnosed with vascular dementia. Supporting evidence for this interpretation comes from a variety of sources, including the results of the control outcome analyses. This would explain the increased risk for the ischaemic heart disease outcome, for which statins are known to be protective, whilst almost no association was observed with the backpain outcome, indicating that most of the uncontrolled confounding is likely to be related to vascular factors. Additionally, we obtained the expected harmful result for Type 2 diabetes, where statins' mechanism of action on this outcome is unlikely to be vascular,[@collins2016a; @macedo2014; @smit2020] Further supporting evidence comes from the increasingly harmful association observed when moving from the probable and possible Alzheimer's disease outcome to the other dementias outcome, and finally to the vascular dementia outcomes. This pattern suggests that the strength of the residual confounding by indication increases as the proportion of cases with a vascular component in an outcome definition increases. A review of other available literature suggests that this observation (a harmful effect of lipid regulating agents on vascular-related outcome due to confounding by indication) is not unusual. Using a conventional epidemiological technique, a previous analysis also found an increased risk of coronary heart disease (analogous to the ischaemic heart disease outcome used in our analysis) in those taking statins (`r estimate(1.31,1.04, 1.66, type = "HR")`).[@danaei2013b] Control for confounding by indication in that study through the use of a trial emulation analysis gave an estimate of 0.89 (95%CI: 0.73-1.09), a comparable though less precise estimate to that observed in RCTs of statin use (`r estimate(0.73,0.67,0.80,type="")`).[@taylor2013]

A secondary limitation is the potential for differential outcome misclassification based on the exposure, as we cannot exclude the possibility that for people with memory complaints, a diagnosis of vascular dementia might be made more frequently than Alzheimer’s disease if their medical records contain prescriptions for lipid-regulating agents. Further, there is also the potential for non-differential misclassification of the outcome ,based on the use of electronic health records to identify dementia cases.[@wilkinson2018; @mcguinness2019validity]

A further limitation stems from uncontrolled confounding due to genetic factors. Number of ApoE $\epsilon4$ alleles represents the strongest genetic risk factor for Alzheimer's disease, but also substantially increases LDL-c levels,[@bennet2007] potentially prompting treatment with a statin or other lipid regulating agent. We were unable to control for ApoE genotype in this analysis as we did not have access to genetic data on participants. As a result, any protective association between LRA use and the Alzheimer's disease outcomes may be masked by residual negative confounding by ApoE.

Finally, as with many studies of dementia, there is a risk of reverse causation in our analysis. Dementia and associated conditions have a long prodromal period, during which preclinical disease could cause indications for the prescription of a lipid-regulating agent.

@@

We have provided new evidence on the association of lipid-regulating agent prescription, predominantly statins, with all-cause dementia, Alzheimer’s disease, vascular dementia, and other dementias. We found limited evidence that use of lipid-regulating agents was associated with incidence of probable or possible Alzheimer’s disease, but they were associated with an increased risk of all-cause, vascular and other dementias. Despite our attempts to account for bias in our analysis, there is a strong potential for residual confounding, misclassification and reverse causation.

This is most likely to explain the unexpected increase in risk of vascular dementia associated with statin use. Future research should aim to address these biases using newer methods such as a trial emulation analysis.

@@

# Declarations

__Ethics approval__

The protocol for this study was approved by the CPRD’s Independent Scientific Advisory Committee (ISAC 15_246R). This study did not directly involve patients, and so further ethical approval was not required.

__Data/code availability__
This analysis used the CPRD-GOLD primary care dataset March 2016 snapshot (ISAC 15_246R), which is available upon application to the CPRD Independent Scientific Advisory Committee. The code lists used to define the outcomes and covariates for this study, in addition to the cleaning and analysis scripts used to create the study cohort and perform the analyses, are available on GitHub (https://github.com/mcguinlu/CPRD-LRA), and were archived at the time of publication on Zenodo (DOI: __TBC__).

__Supplementary data__

Supplementary data are available at IJE online.

# Tables
<!----------------------------------------------------------------------------->
(ref:followUp-caption) Summary of person-time at risk and number of events by exposure group.

@@

# Figures
